Topic modeling on EU and the USA research projects

### 2017-06-24 thesis_report/draft_thesis.tex
Title page & Abstract

\setcounter{secnumdepth}{4}
\setcounter{tocdepth}{4}

\setlength{\parskip}{1em}
\usepackage[justification=centering]{caption}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{algorithm2e}
\usepackage{url}
\usepackage{subfig}
\usepackage{listings}
\usepackage[utf8]{inputenc}
\usepackage{color}

@@

\maketitle
\tableofcontents

\begin{abstract}
This is a simple paragraph at the beginning of the document. 
A brief introduction about the main subject.
\end{abstract}

@@

When formulating the optimization problem in Latent Dirichlet Allocation we 
will use the KL divergence to express the optimization problem. The KL divergence 
is a measure of the difference between two probability distributions $P$ and $Q$. 
It is important to note
that the KL divergence is not symmetrical in $P$ and $Q$ so it’s not a metric.

@@

The Dirichlet distribution is a way to model random probability mass functions
(PMFs)\footnote{If you roll 1000 dice, the theoretical odds of any particular
number showing up (i.e. a 1, 2, 3, 4, 5, or 6) are 1/6. However, you won’t get
that exact distribution in a real experiment due to manufacturing defects. If
you have ten dice, each die will have its own probability mass function (PMF).}
for finite sets. It is also sometimes used as a prior in Bayesian statistics.
The Dirichlet is the multivariate generalization of the beta distribution. It
is an extension of the beta distribution for modeling probabilities for two or
more events; when the result of the event has only 2 values, the Dirichlet
distribution is equal to the beta distribution.

@@

The Dirichlet process is a way to model randomness of a probability mass
function (PMF) with unlimited options (e.g. an unlimited amount of dice in a
bag). The process is similar Polya’s Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}}, only instead of having a set number
of ball colors you have an unlimited amount:

@@

First, we’ll introduce some basic models for topic modeling. The Latent Semantic
Indexing, or LSI, was presented by Deerwester et al. in 1990. The model manages
to deal with the problem that multiple terms can refer to the same meaning,
i.e. synonymy. However, it is not as successful regarding polysemy. The reason
is that every term is represented as just one point in the so-called latent
semantic space. Furthermore, a word that can mean two or more different things
is represented as a weighted average of the different meanings.

@@

This statistical model reflects the intuition that documents exhibit multiple
topics. Each document exhibits the topics with different proportion (step 1);
each word in each document is drawn from one of the topics (step 2b), where the
selected topic is chosen from the per-document distribution over topics (step
2a). All the documents in the corpus share the same set of topics, but each
document exhibits those topics with different proportion - that’s the
distinguishing characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1.

@@

In LDA, our ultimate goal is to find the posterior distribution of latent topic 
variables after observing training data. However, computing this distribution 
is intractable, so we’re forced to approximate. One approximation approach is to 
use an optimization method called Variational Bayes. In the original paper it's 
being used a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

@@

In online Variational Bayes, we only make a single sweep of the entire dataset, 
analyzing a chunk of documents at a time. A chunk could be a single document, 
50 documents, or even the entire dataset. Let’s let a chunk be 1000 documents.

@@

We can see that unlike variational LDA, in online LDA we only need to hold a
small chunk of the data at a time, and once we’re done analyzing it, we never
need it again. As with batch, once we’ve estimated $\lambda$, we can find the 
most probable words for each topic by looking at the word probabilities in 
each row of $\lambda$.\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}

@@

Below, we provide the general topics of the EU projects as listed in the EU 
Open Data Portal \footnote{https://data.europa.eu/euodp/en/data}:
\begin{itemize}
\item Agriculture, Fisheries and Food
	\begin{description}
	\item Agriculture, Food safety, Maritime affairs and fisheries
	\end{description}
\item Business
	\begin{description}
	\item Competition, Enterprise, Single market, Trade
	\end{description}
\item Culture and Education
	\begin{description}
	\item Audiovisual and media, Culture, Education training and youth, Multilingualism
	\end{description}
\item Customs and Tax
	\begin{description}
	\item Customs, Taxation
	\end{description}
\item Development and Humanitarian aid
	\begin{description}
	\item Development and cooperation, Humanitarian aid and Civil protection, Human rights
	\end{description}
\item Economy and Finance
	\begin{description}
	\item Budget, Economic and monetary affairs, Fraud prevention
	\end{description}
\item Employment and Social affairs
\item Enlargement and Foreign affairs
\item Environment and Energy
	\begin{description}
	\item Climate action, Energy, Environment
	\end{description}
\item EU institutions
	\begin{description}
	\item Institutional affairs
	\end{description}
\item Health
	\begin{description}
	\item Health, Sport
	\end{description}
\item Justice and Citizens’ rights
	\begin{description}
	\item EU citizenship, Consumers, Justice and Home affairs
	\end{description}
\item Regions and Local development
	\begin{description}
	\item Regional policy
	\end{description}
\item Science and Technology
	\begin{description}
	\item Space, Digital Economy and Society, Research and Innovation
	\end{description}
\item Transport and Travel
\end{itemize}

@@

To obtain the collection of the documents, we consider as ‘documents’ in LDA the
project objectives, which is an accurate description for the project. We also
added the ‘title’ value of each project to the objectives-documents, to increase
the probability of finding accurate topics. For the majority of the projects the
‘start date’ and ‘end date’ value was null, so we kept the splitting of the
dataset per framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

@@

In general, the STEM (science, technology, engineering and math) related
research and education is becoming more and more popular, which is not
surprising, since the National Science Foundation itself estimates that 80/% of
the jobs available during the next decade will require math and science skills.
Even the concept of STEM was first introduced by Judith A. Ramaley, the former
director of the National Science Foundation’s education and human resources
division.
\begin{center}
\begin{figure}
\includegraphics[width=1.0\textwidth]
{figs/topic-evolution-usa.png}
\caption{Topics over time in USA}
\end{figure}
\end{center}

>>>>

\setcounter{secnumdepth}{4}
\setcounter{tocdepth}{4}
\setlength{\parskip}{1em}

\usepackage[justification=centering]{caption}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{algorithm2e}
\usepackage{url}
\usepackage{subfig}
\usepackage{listings}
\usepackage[utf8]{inputenc}
\usepackage{color}
\usepackage{geometry}

@@

\begin{titlepage}
	\centering
	\includegraphics[width=0.8\textwidth]{figs/logo.jpg}\par\vspace{4cm}
	{\scshape\Large MSc thesis\par}
	\vspace{1.5cm}
	{\Large\bfseries Topic modeling on EU and the USA research projects\par}
	\vspace{2cm}
	{\Large Maria Iliadi\par}
	\vfill
	supervised by\par
	Prof.~Panos Louridas

	\vfill

% Bottom of the page
	{\large \today\par}
\end{titlepage}
\tableofcontents

\begin{abstract}
The purpose of this work is to better identify, quantify and understand themes
in research and innovation over the last 22 years, and to provide valuable
knowledge about the research trends nowadays. A way to approach the issue is
through the projects supported by funding agencies. In particular, public data
are available for the projects funded under the European Union Framework
Programmes, and for project funder by the NSF in the United States. The data
include project descriptions, so it is possible to investigate, based on these
textual data, the research topics that were actually funded over the last two
decades. We first introduce the concept of Latent Dirichlet allocation and its
application in topic modelling in projects funded in the two regions from 1994
to 2017. We will then proceed to show the differences between the two regions in
terms of topic similarity and topic evolution over time. We will discuss both
policy implications, what has been funded over the years, as well as technical
ones, how can one handle such datasets and extract information from them.
\end{abstract}

@@

When formulating the optimization problem in Latent Dirichlet Allocation we 
will use the KL divergence to express the optimization problem. The KL divergence 
is a measure of the difference between two probability distributions $P$ and $Q$. 
It is important to note
that the KL divergence is not symmetrical in $P$ and $Q$ so it's not a metric.

@@

The Dirichlet distribution is a way to model random probability mass functions
(PMFs)\footnote{If you roll 1000 dice, the theoretical odds of any particular
number showing up (i.e. a 1, 2, 3, 4, 5, or 6) are 1/6. However, you won't get
that exact distribution in a real experiment due to manufacturing defects. If
you have ten dice, each die will have its own probability mass function (PMF).}
for finite sets. It is also sometimes used as a prior in Bayesian statistics.
The Dirichlet is the multivariate generalization of the beta distribution. It
is an extension of the beta distribution for modeling probabilities for two or
more events; when the result of the event has only 2 values, the Dirichlet
distribution is equal to the beta distribution.

@@

The Dirichlet process is a way to model randomness of a probability mass
function (PMF) with unlimited options (e.g. an unlimited amount of dice in a
bag). The process is similar Polya's Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}}, only instead of having a set number
of ball colors you have an unlimited amount:

@@

First, we'll introduce some basic models for topic modeling. The Latent Semantic
Indexing, or LSI, was presented by Deerwester et al. in 1990. The model manages
to deal with the problem that multiple terms can refer to the same meaning,
i.e. synonymy. However, it is not as successful regarding polysemy. The reason
is that every term is represented as just one point in the so-called latent
semantic space. Furthermore, a word that can mean two or more different things
is represented as a weighted average of the different meanings.

@@

This statistical model reflects the intuition that documents exhibit multiple
topics. Each document exhibits the topics with different proportion (step 1);
each word in each document is drawn from one of the topics (step 2b), where the
selected topic is chosen from the per-document distribution over topics (step
2a). All the documents in the corpus share the same set of topics, but each
document exhibits those topics with different proportion - that's the
distinguishing characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1.

@@

In LDA, our ultimate goal is to find the posterior distribution of latent topic 
variables after observing training data. However, computing this distribution 
is intractable, so we're forced to approximate. One approximation approach is to 
use an optimization method called Variational Bayes. In the original paper it's 
being used a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

@@

In online Variational Bayes, we only make a single sweep of the entire dataset, 
analyzing a chunk of documents at a time. A chunk could be a single document, 
50 documents, or even the entire dataset. Let's let a chunk be 1000 documents.

@@

We can see that unlike variational LDA, in online LDA we only need to hold a
small chunk of the data at a time, and once we're done analyzing it, we never
need it again. As with batch, once we've estimated $\lambda$, we can find the 
most probable words for each topic by looking at the word probabilities in 
each row of $\lambda$.\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}

@@

Below, we provide the general topics of the EU projects as listed in the EU 
Open Data Portal \footnote{https://data.europa.eu/euodp/en/data}:
\begin{itemize}
\item Agriculture, Fisheries and Food
	\begin{description}
	\item Agriculture, Food safety, Maritime affairs and fisheries
	\end{description}
\item Business
	\begin{description}
	\item Competition, Enterprise, Single market, Trade
	\end{description}
\item Culture and Education
	\begin{description}
	\item Audiovisual and media, Culture, Education training and youth, Multilingualism
	\end{description}
\item Customs and Tax
	\begin{description}
	\item Customs, Taxation
	\end{description}
\item Development and Humanitarian aid
	\begin{description}
	\item Development and cooperation, Humanitarian aid and Civil protection, Human rights
	\end{description}
\item Economy and Finance
	\begin{description}
	\item Budget, Economic and monetary affairs, Fraud prevention
	\end{description}
\item Employment and Social affairs
\item Enlargement and Foreign affairs
\item Environment and Energy
	\begin{description}
	\item Climate action, Energy, Environment
	\end{description}
\item EU institutions
	\begin{description}
	\item Institutional affairs
	\end{description}
\item Health
	\begin{description}
	\item Health, Sport
	\end{description}
\item Justice and Citizens' rights
	\begin{description}
	\item EU citizenship, Consumers, Justice and Home affairs
	\end{description}
\item Regions and Local development
	\begin{description}
	\item Regional policy
	\end{description}
\item Science and Technology
	\begin{description}
	\item Space, Digital Economy and Society, Research and Innovation
	\end{description}
\item Transport and Travel
\end{itemize}

@@

To obtain the collection of the documents, we consider as 'documents' in LDA the
project objectives, which is an accurate description for the project. We also
added the 'title' value of each project to the objectives-documents, to increase
the probability of finding accurate topics. For the majority of the projects the
'start date' and 'end date' value was null, so we kept the splitting of the
dataset per framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

@@

In general, the STEM (science, technology, engineering and math) related
research and education is becoming more and more popular, which is not
surprising, since the National Science Foundation itself estimates that 80/% of
the jobs available during the next decade will require math and science skills.
Even the concept of STEM was first introduced by Judith A. Ramaley, the former
director of the National Science Foundation's education and human resources
division.
\begin{center}
\begin{figure}
\includegraphics[width=1.0\textwidth]
{figs/topic-evolution-usa.png}
\caption{Topics over time in USA}
\end{figure}
\end{center}
